
\begin{acknowledgements}

First and foremost I would like to thank my advisor, Dr. Edward Knightly, whose guidance, patience, and support has been invaluable in producing this thesis and providing me the space to be the best version of myself.  
Additionally, I would like to thank my thesis committee members for their comments, evaluation, and their time spent examining my work. 

I am grateful to my colleagues and especially my collaborators in the Rice ECE department without whom I would have learned very little.
Every Quixote needs their Sancho, and I am thrilled to have been able to tilt on my burro alongside some of the best researchers, dreamers, and friends you can find.
You know who you are and I owe you a drink and more.

I owe a large debt of gratitude to the previous generation of graduate students, particularly the WARP team: Patrick, Chris, Gareth, and Sid, for blazing a path that many of us have followed.
It may have been circuitous, but without it, we would still be lost in the woods.

How any family can put up with ten years of ambivalence is well beyond me, but I love you, Mom \& Dad.
You revealed the beautiful consistency of Science and then let me run double-blind to see what would happen. Well, I became an engineer, but not all is lost!

And finally, to the beautiful woman who beat me to the finish line only to run the race again while holding my hand: Rose, you continue to challenge and inspire me every day.

%The authors would like to thank the following people for their assistance in performing experiments included in this work: Rachel J. Gray (Rice University), Yuqiang Mu (Rice University), and Pablo Salvador (IMDEA Networks Institute). This research was supported by NSF grants CNS-1314822, CNS-1126478, CNS-1012831, and a grant from Cisco Systems, Inc.

\end{acknowledgements}