
%\nobibliography*

\textbf{\textsc{Publications containing work that appears in part in this thesis.
}}
%\noindent\rule{8cm}{0.4pt}

\begin{itemize}
	\item R.~E. Guerra, N.~Anand, C.~W. Shepard, and E.~W. Knightly, ``{Opportunistic
  Channel Estimation for Implicit 802.11af MU-MIMO},'' in \emph{Proceedings of the 25th International Conference on Telecommunications},
  Wurzburg, Germany, Sept. 2016.
	\item C.~W. Shepard, R.~Doost-Mohammady, R.~E. Guerra, and L.~Zhong, ``{ArgosV3: An
  Efficient Many-Antenna Platform},'' in \emph{Proceedings of the 23rd Annual
  International Conference on Mobile Computing and Networking}.\hskip 1em plus
  0.5em minus 0.4em\relax ACM, 2017, pp. 501--503.
	\item C.~W. Shepard, J.~Ding, R.~E. Guerra, and L.~Zhong, ``{Understanding Real
  Many-Antenna MU-MIMO Channels},'' in \emph{Proceedings of IEEE Asilomar Signals,
  Systems, and Computers}, November 2016.
	\item N.~Anand, R.~E. Guerra, and E.~W. Knightly, ``{The Case for UHF-band
  MU-MIMO},'' \emph{Proceedings of ACM MobiCom}, September 2014.
	\item A.~Flores, R.~E. Guerra, E.~W. Knightly, P.~Ecclesine, and S.~Pandey, ``{IEEE
  802.11af: A Standard for TV White Space Spectrum Sharing},'' \emph{IEEE
  Communications Magazine}, October 2013.
	\item C.~Shepard, R.~Doost-Mohammady, J.~Ding, R.~E. Guerra, and L.~Zhong,
  ``Argosnet: A multi-cell many-antenna mu-mimo platform,'' November 2017.
	\item R.~E. Guerra, N.~Anand, and E.~Knightly, ``{Demo: An open-source development
  platform for long-range UHF-connected wifi hotspots},'' \emph{Proc. of ACM
  MobiCom}, September 2014.
	\item R.~E. Guerra, N.~Anand, and E.~W. Knightly, ``{Demo: A Platform for At-Scale
  Wideband UHF MU-MIMO Systems},'' \emph{Proc. of USENIX NSDI}, April 2014.
\end{itemize}


\textbf{\textsc{3rd-party publications utilizing the \ac{WURC} \ac{SDR} platform.}}

%\noindent\rule{8cm}{0.4pt}

\begin{itemize}
	\item T.~Wei, S.~Wang, A.~Zhou, and X.~Zhang, ``Acoustic eavesdropping through
  wireless vibrometry,'' in \emph{Proceedings of the 21st Annual International
  Conference on Mobile Computing and Networking}.\hskip 1em plus 0.5em minus
  0.4em\relax ACM, 2015, pp. 130--141.
	\item S.~Sur and X.~Zhang, ``Bridging link power asymmetry in mobile whitespace
  networks,'' in \emph{Computer Communications (INFOCOM), 2015 IEEE Conference
  on}.\hskip 1em plus 0.5em minus 0.4em\relax IEEE, 2015, pp. 1176--1184.
	\item X.~Zhang and E.~W. Knightly, ``Watch: Wifi in active tv channels,'' \emph{IEEE
  Transactions on Cognitive Communications and Networking}, vol.~2, no.~4, pp.
  330--342, 2016.
	\item C.~A. Tarver, ``Sub-band digital predistortion for noncontiguous carriers:
  Implementation and testing,'' Ph.D. dissertation, Rice University, 2016,
  {Masters Thesis}.
	\item X.~Zhang, ``Watch: Enabling wifi in active tv channels,'' Ph.D. dissertation,
  Rice University, 2014, {Masters Thesis}.
\end{itemize}

